Methods for Projecting Population Growth

### 2018-05-05 final/tex/Paper.tex
Update paper for a4
Change to lstsq
Add south africa

\documentclass{article}
\usepackage[title]{appendix}
\usepackage{authblk}
\usepackage{mathptmx}
\usepackage{url,latexsym,amsmath,amsthm,xspace,rotating,multirow,multicol,xspace,amssymb,paralist}
\usepackage{euscript}
\usepackage{fancybox,xcolor}
\usepackage{longtable}
\usepackage{paralist}
\usepackage[normalem]{ulem}
\usepackage[pdftex]{hyperref}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage{cancel}
\usepackage{mathtools}

@@

The Logistic Growth model is colloquially defined as:

@@

While $N_0$ is trivially obtained from the dataset, $\gamma$ and $K$ are not so simple to obtain. These values are optimized utilizing the least squares method on the dataset, which contains census data since 1910. This produces a model that very closely matches the data, as seen in Figure 1.

\begin{figure}[ht!]
\centering
\includegraphics[width=90mm]{images/lr-graph.png}
\caption{Results of using least squares on logistic growth \label{overflow}}
\end{figure}

@@

\section{Augmented Logistic Growth}

\todo{Fill in details}

@@

\begin{align}
\begin{split}
\label{eq:lv1}
	\frac{dp}{dt} = ap - rpg \\
	\frac{dg}{dt} = rpg - bg
\end{split}
\end{align}
where $p$ is the population of the prey, $g$ is the population of the predators, and $\frac{dp}{dt}$ and $\frac{dg}{dt}$ are their annual population changes. $rpg$ describes the interaction between the two species with $r$ being a regulating coefficient, $a$ being the growth rate of the prey in the absence of predators, and $b$ being the death rate of the predators in the absence of prey. Assuming $g$ and $p$ have the same variation, then the ratio $\frac{g}{p}$ can be evaluated to a constant ${q}$. \ref{eq:lv1} can then be rewritten as:

@@

In order to account for other causes of variation that are not in the predator-prey system, the model can once again be modified to include a function $f(p, g, t)$. This can be seen as a function that modulates the growth of the prey in relation to a theoretical maximum capacity (i.e. due to the lack of natural resources). \ref{eq:lv2} can now be rewritten as:

@@

@@

\subsection{Additional Simulation}

@@

\subsection{Analysis}

>>>>

\documentclass[a4paper]{article}
\usepackage[title]{appendix}
\usepackage{authblk}
\usepackage{mathptmx}
\usepackage{url,latexsym,amsmath,amsthm,xspace,rotating,multirow,multicol,xspace,amssymb,paralist}
\usepackage{euscript}
\usepackage{fancybox,xcolor}
\usepackage{longtable}
\usepackage{paralist}
\usepackage[normalem]{ulem}
\usepackage[pdftex]{hyperref}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage{cancel}
\usepackage{mathtools}
\usepackage{fullpage}
\usepackage[nodayofweek]{datetime}

@@

Perhaps one of the most basic and widely used population growth models is the logistic growth model. It is relatively simple to understand and to adapt to the problem at hand.

As described in class, the logistic growth model is colloquially defined as:

@@

The model indicates that over time, the population grows exponentially, before slowly decreasing in reproduction rate. After this inflection point, the population will begin to even out and eventually grow marginally every year, maintaining a size just under the maximum size $K$.

\subsection{Simulation}

While $N_0$ is trivially obtained from the dataset, $\gamma$ and $K$ must be calculated based on the dataset. These values are optimized utilizing the least squares method. This was used on data collected from the World Bank, which provides information dating back to 1960. The resulting model can be seen in Figure 1.

%\begin{figure}[ht!]
%\centering
%\includegraphics[width=90mm]{images/lr-graph.png}
%\caption{Results of using least squares on logistic growth \label{overflow}}
%\end{figure}

@@

@@

\begin{align}
\begin{split}
\label{eq:lv1}
	\frac{dp}{dt} = ap - rpg \\
	\frac{dg}{dt} = rpg - bg
\end{split}
\end{align}
where $p$ is the population of the prey, $g$ is the population of the predators, and $\frac{dp}{dt}$ and $\frac{dg}{dt}$ are their annual population changes. $rpg$ describes the interaction between the two species with $r$ being a regulating coefficient, $a$ being the growth rate of the prey in the absence of predators, and $b$ being the death rate of the predators in the absence of prey. Assuming $g$ and $p$ have the same variation, then the ratio $\frac{g}{p}$ can be evaluated to a constant ${q}$. Equation \ref{eq:lv1} can then be rewritten as:

@@

In order to account for other causes of variation that are not in the predator-prey system, the model can once again be modified to include a function $f(p, g, t)$. This can be seen as a function that modulates the growth of the prey in relation to a theoretical maximum capacity (i.e. due to the lack of natural resources). Equation \ref{eq:lv2} can now be rewritten as:

@@

These numbers, however, do not make much sense. Simple observation based on the dataset reveals that these parameters result in a massive population increase after the first iteration. The primary culprit, it seems, is their value for $\alpha_3 k_1$. As noted earlier, this value indicates how many additional people are added to the population for every dollar added to GDP. Their value indicates that there are an additional 5.2 people for every dollar, which logically does not make much sense. Potential corrections are noted below, based on additional analysis and simulations.

\subsection{Fixing World Simulation}

Additional simulations were performed both for the world and for various countries. The data was obtained from the UN World Bank dating back to 1960 and the parameters were optimized using both least squares and via minimizing a score function.

Since the paper already performed simulations for the world based on a similar dataset, the result was expected to be fairly similar. However, as indicated earlier, their parameters do not make sense and were likely incorrectly reported. The resulting model indicates a substantially growing population, well past what one could reasonably expect. In order to determine the correct values, both the least squares and scoring minimization methods were used. The least squaring method, unfortunately, resulted in nonsensical results. Therefore, the scoring minimization method was utilized and produced more reasonable results. The parameters were $\alpha_1 =0.00327\%$, $\alpha_3 k_1 = \frac{4.93}{10^6} \frac{Hab.}{US \$}$, $\alpha_2 = \frac{-1.15}{10^{16} US \$}$, $\beta_1 = 7.62\%$, and $\beta_2 = \frac{-3.29}{10^{22} Hab.}$

The resulting graph, while matching the dataset quite nicely, does not make much logical sense. It indicates a very rapid population growth far past 10 billion people before leveling off. This is likely due to the smaller dataset being used - the paper had data going back to 1850 whereas the data used in the simulation only dates back to 1960 due to the lack of accurate population and GDP information dating back much further. However, a big takeaway from this simulation is the similarity of the parameters to the paper's. While the values were clearly different, they were fairly similar and, most notably, $\alpha_3 k_1$ was off by a magnitude of about $10^{-6}$. Based off of this, the original parameters were slightly adjusted to see if a more accurate picture could be obtained. As seen in the figure, the resulting graph is remarkably similar to what the report indicates.

@@

@@

\subsection{United States Simulation}

In order to see how well this model adapts to other situations, additional simulations were performed at the country level. Due to the level of accuracy obtained, the least squares method was used. The first simulation was on the United States. The resulting parameters are $\alpha_1 = 1.07\%$, $\alpha_3 k_1 = \frac{2.76}{10^{7}}\frac{Hab.}{US \$}$, $\alpha_2 = \frac{-1.04}{10^{15} US\$}$, $\beta_1 = 17.9\%$, and $\beta_2 = \frac{-4.63}{10^{10} Hab.}$ The graph can be seen in the figure.

The population is predicted to continue rising steadily until it reaches its peak at just under 375 million people in 2060 before decreasing. Unlike the model for the world population though, this decrease does not appear to level off at an equilibrium. This does not make much logical sense since there is no reasonable explanation for the population to steadily decrease that much based on GDP alone. This d

One point of particular interest is the value for $\beta_1$, which indicates the growth rate of GDP in the absence of other factors. $17.9\%$ is remarkably high, especially when compared to the world as a whole. This is somewhat expected considering how developed the United States is.

\subsection{South Africa Simulation}

The second country simulation was performed with South Africa. This is a good candidate for analysis due to its massive difference to the United States - it was only recently developed and, based off of the data, it has experienced extreme economic hardship in the past several years. Previous simulations ran on data that had a constantly increasing GDP, so this example is used to see how the model performs under more extreme conditions.

The output parameters are $\alpha_1 = 2.22\%$, $\alpha_3 k_1 = \frac{2.77}{10^{6}}\frac{Hab.}{US \$}$, $\alpha_2 = \frac{-8.31}{10^{14} US\$}$, $\beta_1 = 29.0\%$, and $\beta_2 = \frac{-5.21}{10^{9} Hab.}$.
